\textbf{Discovery of a disrupting open cluster in the Galactic halo:\\ a recent star formation event in the leading arm of the Magellanic stream}

### 2018-11-03 ms.tex
gaia section

% \submitjournal{ApJ}
\shorttitle{A recent star formation event in the Magellanic Stream}
\shortauthors{Price-Whelan et al.}

@@

We report the discovery of a young $\tau \sim \clage$, metal-poor $[\textrm{Fe}/\textrm{H}] \sim \clfeh$, low-mass $M \sim 1200~\msun$ stellar association in the Milky Way halo, Galactocentric $(R, z) \sim (23, 15)~\textrm{kpc}$.
The association --- \clustername\ --- is spatially near the leading arm of the gas stream emanating from the Magellanic cloud system, but is located $\approx 60^\circ$ from the Large Magellanic Cloud (LMC) center, on the other side of the Milky Way disk relative to the LMC.
By assuming that the cluster is co-located with \hi gas in the stream, we directly measure the distance to the Magellanic stream, $D \approx \cldist$.
At this location relative to the LMC, the measured distance is inconsistent with predictions from models of the LMC/SMC interaction and infall into the Milky Way that do not account for ram pressure and gas interaction with MW disk.
The estimated age of \clustername\ is consistent with the time of last passage through the Galactic midplane, and we therefore speculate that this star-formation event was triggered by the last disk midplane passage of the Magellanic stream gas, which occurred at a Galactocentric radius $R \approx 30~\kpc$.
However, most details of this idea remain a mystery: the Magellanic Stream is low column density, the Milky Way disk at this large radius has low surface density, and the relative velocity of the stream gas and MW gas is large.
However it formed, the discovery of a young stellar cluster in the Milky Way halo presents an interesting opportunity for study.
\end{abstract}

@@

The halo of the Milky Way is characterized by its old ($\gtrsim 10~\textrm{Gyr}$), metal-poor ($\feh \approx -1.5$) stellar population.
This is understood as a signature of the dominant (in stellar mass) progenitor systems that were accreted and disrupted early on in the formation of the Galaxy: massive dwarf galaxies \citep[e.g.,][]{Deason:2015, Fiorentino:2015}.
These systems likely came in with significant gas reservoirs, but were quenched and stripped through collisional processes that heated and dispersed the gas \citep[e.g.,][]{Mayer:2006}, thus preventing immediate star formation in the deposited gas.
The Milky Way, however, continues to accrete satellite galaxies, as is evidenced by the prominent stellar stream from the Sagittarius dwarf galaxy \citep{Ibata:1994, Majewski:2003}, the presence of the Large and Small Magellanic Clouds (LMC, SMC) and about 50 dwarf satellites within the virial radius of the Galaxy.
While Sagittarius was likely stripped of its neutral gas long ago through repeated passage about the Milky Way \citep{Burton:1999, Tepper-Garcia:2018}, the \lmcsmc\ system is associated with $\approx 8\times 10^8~\msun$ of \hi gas \citep{Bruns:2005}, which extends into leading and trailing gas streams that wrap nearly $\approx 200^\circ$ around the sky \citep{Mathewson:1974, Putman:1998, Bruns:2005, Nidever:2010}.

@@

The origin and formation of the LA features in the MS is still uncertain.
Initial studies of the LA argued that the features closest to the \lmcsmc\ can be traced back to the SMC, thus implying that the gas in the LA was stripped from the SMC \citep{Putman:1998}.
In later work, it has instead been argued that outer features of the LMC appear to lead directly into the LA-1 feature \citep{Nidever:2008}, implying an LMC origin for the leading arm.
Recent chemical abundance measurements along several sightlines passing through the LA again suggested the SMC origin \citep{Fox:2018, Richter:2018}.
Recent hydrodynamical simulations of the MCs on the first infall to the Milky Way, however, showed that the repeated encounters between the MCs strip gas not only from the SMC, but also from the LMC, and this tidally stripped gas creates filamentary gas structures both in the LA and the Magellanic Stream \citep{Pardy:2018}.
Whatever the origin of the gas, it is clear that tidal stripping by the Milky Way is required to form the leading arm \citep{Nidever:2008, Besla:2012}.
However, the LA features deviate from the predicted orbit of the \lmcsmc, implying that ram pressure or interactions with gas in the outer Milky Way disk may have removed orbital energy from the LA gas \citep[e.g.,][]{Bekki:2008}.

@@

In this \articlename, we report the discovery of a clustered association of young stars at the far edge of the LA-2 feature ($L_{\textrm{MS}} \sim 65^\circ$) that is unambiguously located far into the Galactic halo ($d_\odot \sim 30~\textrm{kpc}$) and therefore likely formed from gas in the leading arm of the Magellanic stream as it crossed the Galactic disk.
This provides a clear distance measurement to the leading arm of the MS to be compared with models of \lmcsmc\ interaction and infall into the Milky Way, and also provides an opportunity to study recent star formation in a regime unlike that of the Milky Way. \todo{or any other?}

In \sectionname~\ref{sec:data}, we present the initial discovery with \gaia\ \DR{2} and follow-up observations with DECam to obtain deeper photometry of the region around the association.
In \sectionname~\ref{sec:results}, we derive estimates for the total stellar mass of, age of, and distance to the association, and present a plausible formation scenario.
In \sectionname~\ref{sec:discussion}, we discuss implications for future searches for stars associated with the MS, and discrepancies with existing models of the LA.
We conclude in \sectionname~\ref{sec:conclusion}.

@@

We use astrometric data from the \gaia\ mission (\citealt{Prusti:2016}), data release 2 (\DR{2}; \citealt{Gaia-Collaboration:2018, Lindegren:2018}) to search for distant, comoving multiplets of blue stars.
Our original intent was to search for small, distant, comoving groups of blue horizontal branch stars to identify new candidate satellites of the Milky Way.
We therefore initially select all stars from \gaia\ with parallax $\varpi < 1$, color $-0.5 < (\bprp) < 0$, $G$-band magnitude $G < 20$, and Galactic latitude $|b| > 20^\circ$ (see Appendix~\ref{sec:queries} for database query).
We further exclude stars within a $15^\circ$ radius from the LMC, and a $8^\circ$ radius from the SMC --- 27,895 stars remain after these cuts.
We then cross-match this catalog to itself with both sky positions and proper motions: we search for pairs of stars that have separations $s < 0.5^\circ$ and proper motion differences $|\Delta \mu| < 0.5~\masyr$.
We then combine mutually-connected comoving pairs into small groups of stars that are colocated on the sky and comoving in proper motions, and remove groups that have $<4$ members.
We cross-match the mean sky positions of the groups to locations of local group galaxies \todo{citation} \citep{Mcconaghie} and Milky Way globular clusters \todo{citation} \citep{Harris} and filter out all groups that lie within 1 degree of these known objects.
After these filters, just one group of blue, comoving stars remains at $(\textrm{RA}, \textrm{Dec}) \sim (179, -29)^\circ$.

We then query all objects from the \gaia\ \DR{2} catalog within a rectangle centered on the nominal position of this group, with a width of $12^\circ$ and a height of $12^\circ$ in the equatorial (ICRS) coordinate system (see Appendix~\ref{sec:queries} for the database query).
We cross-match the resulting catalog with \acronym{GALEX} \citep{Martin:2005} using a positional cross-match requiring source separation $<2~\textrm{arcsec}$, and with \pans\ \citep{Chambers:2016} using the cross-match provided by the \gaia\ science archive.
\figurename~\ref{fig:cmds} shows the \gaia\ color-magnitude diagram (left) and the \acronym{GALEX} NUV minus \pans\ $g$-band versus \gaia\ \bprp\ color-color diagram (right) from the cross-matched photometry.
All magnitudes are corrected for dust extinction using the SFD \citep{Schlegel:1998, Sclafly:2011} dust map.
The green polygon in the left panel of \figurename~\ref{fig:cmds} highlights the over-density of blue stars that were identified as comoving from the proper motion search described above: surprisingly, this appears to be a young, distant main sequence, rather than an old population of horizontal branch stars.
The lines in the right panel of \figurename~\ref{fig:cmds} show isochrones from the \acronym{MIST} \citep[blue;][]{Choi:2016} and \acronym{PARSEC} \citep[red;][]{Bressan:2012, Marigo:2017} isochrone libraries for a 50 Myr stellar population with metallicity $\feh = -0.5$ (dashed) or $\feh = -1.5$ (solid). \todo{Remake this plot with just FeH=-1.}

% Notebook: Figure-CMDs
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/gaia-galex-cmd.pdf}
\caption{Color-magnitude and color-color diagrams with \gaia\ ($G$, BP, RP), \acronym{GALEX} (NUV), and \pans\ ($g$) magnitudes.
Shaded pixels (orange) show regions of high density and points (black markers) are plotted when there are fewer than 8 stars per pixel.
Green polygon in left panel highlights the over-density of blue, comoving stars identified by the search described in \sectionname~\ref{sec:discovery}.
Red and blue lines in right panel show the zero-age main sequence of a $50~\textrm{Myr}$ stellar population with the metallicity indicated from the \acronym{MIST} \citep{TODO} and \acronym{PARSEC} \citep{TODO} isochrone libraries.
}
\label{fig:cmds}
\end{figure}

@@

\subsection{Relation to the Magellanic Stream}
\label{sec:higas}

At the sky location and distance of \clustername, i.e. well into the Galactic halo, the only plausible gas reservoirs that could have formed a young cluster are the Magellanic Stream, or a previously unknown high velocity cloud (HVC).
HVCs are thought to either be accreted and therefore lower metallicity than typical present-day Milky Way gas, or ejected from the Milky Way through a ``Galactic fountain''-like process and therefore comparable metallicity to disk gas.
Both processes clearly occur: the Magellanic Stream itself is evidence of gas accretion in to the Milky Way halo, and the well-known Smith Cloud \citep{TODO} is a metal-rich ($\feh \sim 0.5$), HVC that could have originated from the Galactic disk \citep{TODO}.

Given the low metallicity of the stars in \clustername, the gas it formed from was likely extragalactic, as any violent star-forming regions in the Milky Way disk have, at present-day, significantly higher metallicity than this cluster.
\figurename~\ref{fig:XXXX} shows the \hi column density in the region around the leading arm of the Magellanic Stream, with the location of \clustername\ marked.
While no bulk metallicity measurements exist for the LA gas, \todo{...recent measurements suggest comparable to SMC?} \citep{Fox:2018}.
Given its extragalactic origin, its proximity to the Magellanic Stream suggests that \clustername\ could have formed from this gas.
Previously studies have detected young stars in the Leading Arm and gas in the periphery of the Magellanic Clouds \citep{Casetti-Dinescu:2014, MoniBidin:2017}, however, this is the first time that an entire star cluster has been detected so far from the Clouds.

@@

% We evaluate the hypothesis that the cluster formed in the Leading Arm.  The mean proper motion values in the Magellanic Stream coordinate system \citep{Nidever:2008} are ($\mu_{\rm MSL}$,$\mu_{\rm MSB}$)=($+$1.X,0.X) mas yr$^{-1}$ which is very similar to the mean values of the LMC and SMC with ($+$1.X,0.X) mas yr$^{-1}$ and ($+$1.X,0.X) mas yr$^{-1}$, respectively.  The simulations of \citet{Besla:2012} also predict proper motions at this position of approximately ($+$1.X,0.X) mas yr$^{-1}$.  The Magellanic Stream simulations give distances of the Leading Arm at this position of $\sim$50 kpc, however, many of these do not include the effects of ram pressure from the hot MW halo which could cause the gas to drop in its orbit to smaller Galactocentric distances (NEED TO DOUBLE-CHECK THIS).  In addition Figure \ref{fig_gass} shows that there is Leading Arm \hi~gas at the position of the position of the cluster with a velocity of \vlsr=70 \kmse.

@@

Query to retrieve \gaia\ data around the blue, comoving group found and discussed in \sectionname~\ref{sec:data}:
\begin{verbatim}
SELECT *
FROM gaiadr2.gaia_source
WHERE (parallax < 1 OR parallax IS NULL)
    AND ra > 173.5 AND ra < 185.5
    AND dec > -35.0 AND dec < -23.0
\end{verbatim}

>>>>

% \submitjournal{ApJ}
\shorttitle{A recent star formation event in the Magellanic stream}
\shortauthors{Price-Whelan et al.}

@@

We report the discovery of a young $\tau \sim \clage$, metal-poor $[\textrm{Fe}/\textrm{H}] \sim \clfeh$, low-mass $M \sim 1200~\msun$ stellar association in the Milky Way halo, Galactocentric $(R, z) \sim (23, 15)~\textrm{kpc}$.
The association --- \clustername\ --- is spatially near the leading arm of the gas stream emanating from the Magellanic cloud system, but is located $\approx 60^\circ$ from the Large Magellanic Cloud (LMC) center, on the other side of the Milky Way disk relative to the LMC.
By assuming that the cluster is co-located with \hi gas in the stream, we directly measure the distance to the Magellanic stream, $D \approx \cldist$.
At this location relative to the LMC, the measured distance is inconsistent with predictions from models of the LMC/SMC interaction and infall into the Milky Way that do not account for ram pressure and gas interaction with MW disk.
The estimated age of \clustername\ is consistent with the time of last passage through the Galactic midplane, and we therefore speculate that this star-formation event was triggered by the last disk midplane passage of the Magellanic stream gas, which occurred at a Galactocentric radius $R \approx 30~\kpc$.
However, most details of this idea remain a mystery: the Magellanic stream is low column density, the Milky Way disk at this large radius has low surface density, and the relative velocity of the stream gas and MW gas is large.
However it formed, the discovery of a young stellar cluster in the Milky Way halo presents an interesting opportunity for study.
\end{abstract}

@@

The halo of the Milky Way is characterized by its old ($\gtrsim 10~\textrm{Gyr}$), metal-poor ($\feh \approx -1.5$) stellar population.
This is understood as a signature of the dominant (in stellar mass) progenitor systems that were accreted and disrupted early on in the formation of the Galaxy: massive dwarf galaxies \citep[e.g.,][]{Deason:2015, Fiorentino:2015}.
These systems likely came in with significant gas reservoirs, but were quenched and stripped through collisional processes that heated and dispersed the gas \citep[e.g.,][]{Mayer:2006}, thus preventing immediate star formation in the deposited gas.
The Milky Way, however, continues to accrete satellite galaxies, as is evidenced by the prominent stellar stream from the Sagittarius dwarf galaxy \citep{Ibata:1994, Majewski:2003}, the presence of the Large and Small Magellanic Clouds (LMC, SMC), and about 50 dwarf satellites within the virial radius of the Galaxy.
While Sagittarius was likely stripped of its neutral gas long ago through repeated passage about the Milky Way \citep{Burton:1999, Tepper-Garcia:2018}, the \lmcsmc\ system is associated with $\approx 8\times 10^8~\msun$ of \hi gas \citep{Bruns:2005}, which extends into leading and trailing gas streams that wrap nearly $\approx 200^\circ$ around the sky \citep{Mathewson:1974, Putman:1998, Bruns:2005, Nidever:2010}.

@@

The origin and formation of the LA features in the MS is still uncertain.
Initial studies of the LA argued that the features closest to the \lmcsmc\ can be traced back to the SMC, thus implying that the gas in the LA was stripped from the SMC \citep{Putman:1998}.
In later work, it has instead been argued that outer features of the LMC appear to lead directly into the LA-1 feature \citep{Nidever:2008}, implying an LMC origin for the leading arm.
Recent chemical abundance measurements along several sightlines passing through the LA again suggested the SMC origin \citep{Fox:2018, Richter:2018}.
Recent hydrodynamical simulations of the MCs on the first infall to the Milky Way, however, showed that the repeated encounters between the MCs strip gas not only from the SMC, but also from the LMC, and this tidally stripped gas creates filamentary gas structures both in the LA and the Magellanic stream \citep{Pardy:2018}.
Whatever the origin of the gas, it is clear that tidal stripping by the Milky Way is required to form the leading arm \citep{Nidever:2008, Besla:2012}.
However, the LA features deviate from the predicted orbit of the \lmcsmc, implying that ram pressure or interactions with gas in the outer Milky Way disk may have removed orbital energy from the LA gas \citep[e.g.,][]{Bekki:2008}.

@@

In this \articlename, we report the discovery of a clustered association of young stars at the far edge of the LA-2 feature ($L_{\textrm{MS}} \sim 65^\circ$) that is unambiguously located far into the Galactic halo ($D \sim \cldist$) and therefore plausibly formed from gas in the leading arm of the Magellanic stream as it crossed the Galactic disk.
This provides a clear distance measurement to the leading arm of the MS to be compared with models of \lmcsmc\ interaction and infall into the Milky Way, and also provides an opportunity to study recent star formation in a regime unlike that of the Milky Way.
The discovery of \clustername\ will enable new modeling efforts that track the infall of the LMC-SMC, the tidal stripping of Magellanic gas, and the interaction of this gas with the Milky Way.

In \sectionname~\ref{sec:data}, we present the initial discovery with \gaia\ \DR{2} and follow-up observations with \decam\ to obtain deeper photometry of the region around the association.
In \sectionname~\ref{sec:results}, we use the \decam\ photometry to infer the age, metallicity, and distance to \clustername, and discuss plausible formation scenarios for this cluster.
% In \sectionname~\ref{sec:discussion}, we discuss implications for future searches for stars associated with the MS, and discrepancies with existing models of the LA.
We conclude in \sectionname~\ref{sec:conclusion}.

@@

We use astrometric data from the \gaia\ mission (\citealt{Prusti:2016}), data release 2 (\DR{2}; \citealt{Gaia-Collaboration:2018, Lindegren:2018}) to search for distant, comoving multiplets of blue stars.
Our original intent was to search for small, distant, comoving groups of blue horizontal branch stars to identify new candidate satellites of the Milky Way.
We therefore initially select all stars from \gaia\ with parallax $\varpi < 1$, color $-0.5 < (\bprp) < 0$, $G$-band magnitude $G < 20$, and Galactic latitude $|b| > 20^\circ$ (see Appendix~\ref{sec:queries} for database query).
We further exclude stars within a $15^\circ$ radius from the LMC, and a $8^\circ$ radius from the SMC --- 27,895 stars remain after these cuts.
We then cross-match this catalog to itself with both sky positions and proper motions: we search for pairs of stars that have separations $s < 0.5^\circ$ and proper motion differences $|\Delta \mu| < 0.5~\masyr$.
We then combine mutually-connected comoving pairs into small groups of stars that are colocated on the sky and comoving in proper motions, and remove groups that have $<4$ members.
We cross-match the mean sky positions of the groups to locations of local group galaxies \citep{McConnachie:2012} and Milky Way globular clusters \citep[2010 edition;][]{Harris} and filter out all groups that lie within 1 degree of these known objects.
After these filters, just one group of blue, comoving stars remains at $(\textrm{RA}, \textrm{Dec}) \sim (179, -29)^\circ$.

We then query all objects from the \gaia\ \DR{2} catalog within a rectangle centered on the nominal position of this group, with a width of $5^\circ$ and a height of $5^\circ$ in the equatorial (ICRS) coordinate system (see Appendix~\ref{sec:queries} for the database query).
\figurename~\ref{fig:cmds} shows the \gaia\ data for this region.
The left panel of \figurename~\ref{fig:cmds} shows the \gaia\ color-magnitude diagram, with the blue over-density highlighted by the polygon (blue) and under-plotted with a $100~\textrm{Myr}$, $\feh = -1.1$ \acronym{MIST} isochrone \citep[red line;][]{Dotter:2016, Choi:2016, Paxton:2011, Paxton:2013, Paxton:2015}: Surprisingly, this appears to be a young, distant main sequence, rather than an old population of horizontal branch stars.
The middle and right panels of \figurename~\ref{fig:cmds} show sky positions and proper motions, respectively, of all stars in this sky region (grey background density), and only stars in the CMD selection polygon as markers (blue).

The \gaia\ data demonstrate the presence of a co-moving, clustered over-density of blue stars, but the \gaia\ photometry is too shallow to resolve anything but the brightest main sequence stars.
With the \gaia\ data alone, the distance, age, and metallicity of the cluster therefore cannot be determined, as these quantities are degenerate where the main sequence is nearly vertical.
In the next section, we describe deeper \decam\ imaging obtained over a portion of the cluster.

% Notebook: Figure-CMDs
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/gaia-cmd-pm.pdf}
\caption{Color-magnitude diagram, sky positions, and proper motions from \gaia\ \DR{2} for the region around $(\alpha, \delta) \sim (179, -29)^\circ$.
\textbf{Left panel}: \gaia\ color-magnitude diagram (CMD) for all sources with $G < 20$ and sky position within 3 degrees of $(\alpha, \delta) = (179.5, -28.8)^\circ$.
Red line shows a $100~\textrm{Myr}$, $\feh = -1.1$ \acronym{MIST} isochrone shifted to a distance of $30~\kpc$.
Shaded pixels (orange) show regions of high density and points (black markers) are plotted when there are fewer than 8 stars per pixel.
\textbf{Middle panel}: Sky positions for all sources in the shown sky region (grey 2D histogram).
Blue markers show only sources in the blue selection polygon shown in the CMD (left panel).
\textbf{Right panel}: Proper motions for all sources in the shown sky region (grey 2D histogram).
Blue markers again show only sources in the blue CMD selection region (left panel).
}
\label{fig:cmds}
\end{figure}

@@

\subsection{Relation to the Magellanic stream}
\label{sec:higas}

At the sky location and distance of \clustername, i.e. well into the Galactic halo, the only plausible gas reservoirs that could have formed a young cluster are the Magellanic stream, or a previously unknown high velocity cloud (HVC).
HVCs are thought to either be accreted and therefore lower metallicity than typical present-day Milky Way gas, or ejected from the Milky Way through a ``Galactic fountain''-like process and therefore comparable metallicity to disk gas.
Both processes clearly occur: the Magellanic stream itself is evidence of gas accretion in to the Milky Way halo, and the well-known Smith Cloud \citep{TODO} is a metal-rich ($\feh \sim 0.5$) HVC that could have originated from the Galactic disk \citep{TODO}.
Given the low metallicity of the stars in \clustername, the gas it formed from was likely extragalactic, as any violent star-forming regions in the Milky Way disk that could have driven gas out into the halo have, at present-day, significantly higher metallicity than this cluster.

Here we take a closer look at the \hi gas in the vicinity of \clustername.
\figurename~\ref{fig:XXXX} shows the \hi column density in the region around the leading arm of the Magellanic stream, with the location of \clustername\ marked.
While no bulk metallicity measurements exist for the LA gas, \todo{...recent measurements suggest comparable to SMC?} \citep{Fox:2018}.
Given its extragalactic origin, its proximity to the Magellanic stream suggests that \clustername\ could have formed from this gas.
Previously studies have detected young stars in the Leading Arm and gas in the periphery of the Magellanic Clouds \citep{Casetti-Dinescu:2014, MoniBidin:2017}, however, this is the first time that an entire star cluster has been detected so far from the Clouds.

@@

% We evaluate the hypothesis that the cluster formed in the Leading Arm.  The mean proper motion values in the Magellanic stream coordinate system \citep{Nidever:2008} are ($\mu_{\rm MSL}$,$\mu_{\rm MSB}$)=($+$1.X,0.X) mas yr$^{-1}$ which is very similar to the mean values of the LMC and SMC with ($+$1.X,0.X) mas yr$^{-1}$ and ($+$1.X,0.X) mas yr$^{-1}$, respectively.  The simulations of \citet{Besla:2012} also predict proper motions at this position of approximately ($+$1.X,0.X) mas yr$^{-1}$.  The Magellanic stream simulations give distances of the Leading Arm at this position of $\sim$50 kpc, however, many of these do not include the effects of ram pressure from the hot MW halo which could cause the gas to drop in its orbit to smaller Galactocentric distances (NEED TO DOUBLE-CHECK THIS).  In addition Figure \ref{fig_gass} shows that there is Leading Arm \hi~gas at the position of the position of the cluster with a velocity of \vlsr=70 \kmse.

@@

Query to retrieve \gaia\ data around the blue, comoving group found and discussed in \sectionname~\ref{sec:data}:
\begin{verbatim}
SELECT *
FROM gaiadr2.gaia_source
WHERE (parallax < 1 OR parallax IS NULL)
    AND ra > 177 AND ra < 182
    AND dec > -31.3 AND dec < -26.3
\end{verbatim}
